| Base de datos de abejas ibéricas \n| Iberian bees database\n

### 2022-07-02 Manuscript/manuscript_new_figure.Rmd
new figure

Here we have compiled an Iberian bee database of occurrence records. For this, we have digitalize natural history records from museums, compiled more than 100 datasets contributed by individual researchers, and gathered publicly available information from online repositories. After data cleaning and harmonization, we release the first `r formatC(nrow(data),format="d", big.mark=",")` records from the Iberian Peninsula and the Balearic Islands. **Figure 1** shows the spatial distribution of records, which as expected it is biased to the main urban areas. Overall, the database contains information of `r length(unique(data$Accepted_name))` different species and in **Figure 2** we show the 20 most frequently collected species. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting bee species (**Fig. 3A**). As expected, most records were recorded in spring with May being the month with highest number of records (**Fig. 3B**). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where bee specimens were collected. Here, we release the version 1.0 of the Iberian bee database with name "IberianBees". Although this database is a work in progress, we believe that this is a good starting point to better understand and conserve bee biodiversity. It can be accessed at the following permanent link: <https://doi.org/10.5281/zenodo.6354502>.

@@

En este estudio hemos creado una base datos de abejas ibéricas mediante la unificación de: (i) los esfuerzos de digitalización de museos, (ii) conjuntos de datos de investigadores, e (iii) información de repositorios en línea de libre acceso. Tras la depuración y homogeneización de los datos, se han liberado los primeros `r formatC(nrow(data),format="d", big.mark=",")` registros que abarcan la península ibérica e Islas Baleares. La **Figura 1** muestra la distribución espacial de los registros, observándose, tal y como se esperaba, un sesgo hacia ciertas áreas cercanas a los grandes núcleos urbanos. Esta base de datos contiene información sobre `r length(unique(data$Accepted_name))` diferentes especies y el número de obersevaciones de las 20 más comunes se encuentran representadas en la **Figura 2**. Los datos han sido recopilados entre los años `r min(data$Year, na.rm = T)` y `r max(data$Year, na.rm = T)`, con un ascendente número de registros en el tiempo que refleja el creciente interés de documentar a este grupo taxonómico (**Fig. 3A**). Además, como es de esperar, la mayoría de registros tienen lugar en los meses de primavera, siendo mayo el mes que cuenta con un mayor número de registros (**Fig. 3B**). El `r round(geo/nrow(data),2)*100`% de estos registros presenta información georreferenciada y el resto de registros al menos contiene información de la localidad del evento de captura u observación del espécimen. También un `r round(date/nrow(data),2)*100`% de los registros aporta la fecha de captura, y un `r round(flw/nrow(data),2)*100`% tiene información de la planta donde se capturó u observó el espécimen. Esta es la versión 1.0 de la base de datos de abejas ibéricas con nombre "IberianBees" que, aunque se seguirá actualizando, creemos que es un buen punto de partida para mejorar nuestro conocimiento y contribuir a la conservación de este grupo taxonómico tan importante para nuestros ecosistemas y la producción agrícola. Se puede acceder desde el siguiente link permanente: <https://doi.org/10.5281/zenodo.6354502>.

@@

**Figure 1**. Heatmap of the number of bee records on the Iberian Peninsula. The records are in logarithmic scale with a gradient of colours. Areas with high number of records are coloured in yellow and areas with low number of records in purple.

**Figure 2**. Barplot of the number of records of the 20 most common species.

**Figure 3**. Number of records in the database by year (A) and month (B).

@@

**Figura 1**. Mapa de calor del número de registros de abejas en la península ibérica. Los registros se encuentran en escala logarítmica y se muestran con un gradiente de color. Zonas con alta densidad de registros aparecen en amarillo y zonas con una baja densidad en morado.

**Figura 2**. Gráfico de barras del número de registros de las 20 especies más comunes.

**Figura 3**. Número de registros en la base de datos por año (A) y por mes (B).

@@

@@

>>>>

Here we have compiled an Iberian bee database of occurrence records. For this, we have digitalize natural history records from museums, compiled more than 100 datasets contributed by individual researchers, and gathered publicly available information from online repositories (see **Fig. 1** for workflow). After data cleaning and harmonization, we release the first `r formatC(nrow(data),format="d", big.mark=",")` records from the Iberian Peninsula and the Balearic Islands. **Figure 2** shows the spatial distribution of records, which as expected it is biased to the main urban areas. Overall, the database contains information of `r length(unique(data$Accepted_name))` different species and in **Figure 3** we show the 20 most frequently collected species. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting bee species (**Fig. 4A**). As expected, most records were recorded in spring with May being the month with highest number of records (**Fig. 4B**). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where bee specimens were collected. Here, we release the version 1.0 of the Iberian bee database with name "IberianBees". Although this database is a work in progress, we believe that this is a good starting point to better understand and conserve bee biodiversity. It can be accessed at the following permanent link: <https://doi.org/10.5281/zenodo.6354502>.

@@

En este estudio hemos creado una base datos de abejas ibéricas mediante la unificación de: (i) los esfuerzos de digitalización de museos, (ii) conjuntos de datos de investigadores, e (iii) información de repositorios en línea de libre acceso (ver **Fig. 1** para ver flujo de trabajo). Tras la depuración y homogeneización de los datos, se han liberado los primeros `r formatC(nrow(data),format="d", big.mark=",")` registros que abarcan la península ibérica e Islas Baleares. La **Figura 2** muestra la distribución espacial de los registros, observándose, tal y como se esperaba, un sesgo hacia ciertas áreas cercanas a los grandes núcleos urbanos. Esta base de datos contiene información sobre `r length(unique(data$Accepted_name))` diferentes especies y el número de obersevaciones de las 20 más comunes se encuentran representadas en la **Figura 3**. Los datos han sido recopilados entre los años `r min(data$Year, na.rm = T)` y `r max(data$Year, na.rm = T)`, con un ascendente número de registros en el tiempo que refleja el creciente interés de documentar a este grupo taxonómico (**Fig. 4A**). Además, como es de esperar, la mayoría de registros tienen lugar en los meses de primavera, siendo mayo el mes que cuenta con un mayor número de registros (**Fig. 4B**). El `r round(geo/nrow(data),2)*100`% de estos registros presenta información georreferenciada y el resto de registros al menos contiene información de la localidad del evento de captura u observación del espécimen. También un `r round(date/nrow(data),2)*100`% de los registros aporta la fecha de captura, y un `r round(flw/nrow(data),2)*100`% tiene información de la planta donde se capturó u observó el espécimen. Esta es la versión 1.0 de la base de datos de abejas ibéricas con nombre "IberianBees" que, aunque se seguirá actualizando, creemos que es un buen punto de partida para mejorar nuestro conocimiento y contribuir a la conservación de este grupo taxonómico tan importante para nuestros ecosistemas y la producción agrícola. Se puede acceder desde el siguiente link permanente: <https://doi.org/10.5281/zenodo.6354502>.

@@

**Figure 1** Workflow to build the 'IberianBees' database.

**Figure 2**. Heatmap of the number of bee records on the Iberian Peninsula. The records are in logarithmic scale with a gradient of colours. Areas with high number of records are coloured in yellow and areas with low number of records in purple.

**Figure 3**. Barplot of the number of records of the 20 most common species.

**Figure 4**. Number of records in the database by year (A) and month (B).

@@

**Figure 1** Fujo de trabajo para construir la base de datos de 'Iberianbeess'.

**Figura 2**. Mapa de calor del número de registros de abejas en la península ibérica. Los registros se encuentran en escala logarítmica y se muestran con un gradiente de color. Zonas con alta densidad de registros aparecen en amarillo y zonas con una baja densidad en morado.

**Figura 3**. Gráfico de barras del número de registros de las 20 especies más comunes.

**Figura 4**. Número de registros en la base de datos por año (A) y por mes (B).

@@

```{r, out.width="100%", include=TRUE, fig.align="center", echo=FALSE}
knitr::include_graphics("workflow.pdf")
```

@@

# Figura 4
